---
title: "04_describe"
format: html
editor: visual
bibliography: references.bib
---

```{r}
library(tidyverse)
```

## Lets compile general data description here

```{r}
# load in augmented data for description
data = read_tsv('../data/augmented_meta_gene_table.tsv')
```

Hidradenitis suppurativa (HS) is a severe, debilitating chronic inflammatory disease affecting hair follicles with a predilection for intertriginous areas.[@freudenberg2023]. We looked at this study of 20 HS patients at the University of North Carolina Department of Dermatology clinics from October 2017 through February 2018. Whole skin punch biopsies were collected from a lesional site and a matching non-lesional site located 5 cm from the site of inflammation for all the patients[@freudenberg2023]. We want to look at the genes that are significantly differentially expressed between the lesional and perilesional sites; the idea is the genes that are significantly over/under-expressed in the lesional sites might be involved in molecular pathways that affect HS.

```{r}
meta_cols <- c('Sample',
               'Gender',
               'Race',
               'Smoker',
               'Age',
               'BMI')

```

```{r}

# Data description like number of genes, number of patients etc.
data |> 
  select(gene) |> 
  unique() |> 
  nrow()
```

There are 304 genes we are looking at across the lesional and perilesional tissue samples.

```{r}
wide_data <- 
  data |>
  select(all_of(meta_cols),
         gene,
         Lesion,
         Perilesion) |> 
  pivot_longer(c(Lesion, Perilesion),
               names_to = 'Tissue_type',
               values_to = 'expression') |> 
  pivot_wider(names_from = gene, 
              values_from = expression)

wide_data |> 
  filter(Tissue_type == 'Perilesion') |> 
  group_by(Gender) |> 
  count() |> 
  ungroup()
```

Out of the 20 patients selected for this study, 17 were females and 3 were males.

```{r}
wide_data |> 
  group_by(Tissue_type) |>
  count() |> 
  ungroup()
```

20 lesional and perilesional samples each were collected from the patients, i.e, a pair of lesional and perilesional samples from each patient.

```{r}
# Change to this one
wide_data |>
  group_by(Smoker, Gender) |>
  summarize(n = n()) |> 
  ungroup()
```

6 patients were ex-smokers, 9 were non-smokers and 5 were smokers.

5 female patients were ex-smokers, 9 were non-smokers and 3 were smokers.

1 male patient was an ex-smoker, and 2 were smokers. So no male patient was a non-smoker.

```{r}
attribute_dist_table <- wide_data |> 
  select(Sample, Age, Gender, BMI, Smoker, Race)

attribute_dist_table
```

```{r}
# Creating agegroup

attribute_dist_table <- 
  attribute_dist_table |>  
  mutate(Age_group = case_when(
    Age >= 15 & Age <= 20 ~ '[15:20]',
    Age > 20 & Age <= 30 ~ '(20:30]',
    Age > 30 & Age <= 40 ~ '(30:40]',
    Age > 40 & Age <= 50 ~ '(40:50]',
    Age > 50 & Age <= 60 ~ '(50:60]'
  )) 
```

```{r}
# Write out catagories
attribute_dist_table <- 
  attribute_dist_table |> 
  mutate(Gender = if_else(Gender == 'F',
                          'Female',
                          'Male')
         ) |>
  mutate(Smoker = case_when(Smoker == 'E'~'Ex-smoker',
                            Smoker == 'Y'~'Yes',
                            Smoker == 'N'~'No')
         )
```

```{r}
attribute_dist_table |>
  ggplot(aes(x = Age_group,
             fill = Gender
             )
         ) +
  geom_bar(color = "black", 
           alpha = 0.5) +
  geom_hline(yintercept = 0,
             color = 'black',
             size = 0.5) +
  facet_wrap(~Gender) +
  labs(title = 'Age Distribution by Gender',
       x = 'Age group',
       y = 'n') +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        plot.title = element_text(hjust = 0.5,
                                  vjust = 2,
                                  face = 'bold'),
        strip.text = element_text(size = 11), # Adjust facet title size
        ) +
  guides(fill = "none") # Remove legend

```

The age distribution of the patients is shown above.

```{r}
attribute_dist_table |>
  ggplot(aes(x = Age_group,
             fill = Gender
             )
         ) +
  geom_bar(color = "black", 
           alpha = 0.5) +
  geom_hline(yintercept = 0,
             color = 'black',
             size = 0.5) +
  facet_wrap(~Smoker) +
  labs(title = 'Age Distribution by Smoking history and Gender',
       x = 'Age group',
       y = 'n') +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        plot.title = element_text(hjust = 0.5,
                                  vjust = 2,
                                  face = 'bold'),
        strip.text = element_text(size = 11),
        axis.text.x = element_text(angle = 45,
                                   hjust = 1,
                                   vjust = 1.5)
        )
```

```{r}
attribute_dist_table |> 
  ggplot(aes(x = BMI)) +
  geom_histogram(fill = "salmon", color = "black", alpha = 0.5, bins = 10) +
  facet_wrap(~Smoker) +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

The BMI distribution of the patients is shown above.

```{r}
attribute_dist_table |> 
  ggplot(aes(x = BMI)) +
  geom_histogram(fill = "salmon", color = "black", alpha = 0.5, bins = 10) +
  facet_wrap(~Race) +
  theme_minimal() +
  theme(panel.grid = element_blank())
```
